A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-11-09 analysis/paper/paper.Rmd
Changes after peer review

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed 'shoreline dating' which is widely applied. However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located below the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. Monte Carlo simulation is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous hypotheses that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

keywords: |
  Shoreline dating; Stone Age; Settlement patterns; Scandinavia; Relative sea-level change
---

@@

```{r libraries, eval = TRUE, results = 'hide'}
library(here)
library(tidyverse)
library(sf)
library(ggridges)
library(ggthemes)
library(ggnewscale)
library(patchwork)
library(terra)
library(raster)
library(oxcAAR)
library(vapour)
library(topoDistance)
library(IRanges)
library(stringr)
library(kableExtra)
quickSetupOxcal()

@@

# Load processed site and radiocarbon data, and load functions
load(here("analysis/data/derived_data/01data.RData"))
load(here("analysis/data/derived_data/02data.RData"))
load(here("analysis/data/derived_data/06data.RData"))
load(here("analysis/data/derived_data/08agediff.RData"))
source(here("analysis/script/04analysis_functions.R"), local = knitr::knit_global())

# Exclude Lunaveien and Dybdalshei 2 (See supplementary material)
sites_sa <- filter(sites_sa, !(name %in% c("Dybdalshei 2", "Lunaveien")))

@@

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit (the highest elevation of the sea after the retreat of the ice) were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [see e.g. @berg-hansen2009; @eskeland2017; @nummedal1923; @simpson2009]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].

@@

Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @akerlund1995; @åstveit2018; @solheim2020; see also @bjerck2008a; @kleppe1985; @ramstad2009]. This approach has a couple of limitations. First, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g. @åstveit2018; @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that the conclusion supports the hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.

@@

The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase. The isobases are here contours indicating equal shoreline displacement over the same time span [@svendsen1987, 116]. To minimise the impact of variable uplift rates, the cored basins are located in as constrained of an area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)). Furthermore, as the uplift has been greater immediately following the retreat of the ice, such adjustments, and thus potential uncertainty, will be more critical further back in time. The resulting SLIPs are most commonly interpreted as representing the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range. For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [30cm at the standard port Helgeroa in Larvik, @mapping2021]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs as they are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g. @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

The archaeological data compiled for the analysis consists of excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number `r nrow(sites)` sites of which `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates. Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon date ranges that intersect the Stone Age (9500--1700 BCE), with 95% probability. These sites and ^14^C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History of the University of Oslo---the institution responsible for archaeological excavations and data curation in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

@@

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of RSL change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)). This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDistance package for R [@wang2019]. The topographic distance was measured between the site polygon and the horizontally closest point on the shoreline. This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline. Not finding the topographically closest point significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results given the distances considered. The shortest topographic path was found using the Moore neighbourhood of eight cells [e.g. @conolly2006, 253; @herzog2013]. In the case where the sea polygons intersect the site polygon, all distance measures were set to zero. In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material [cf. @crema2010, 1125]. Hovland 5 was chosen for this evaluation as it has an imprecise age and is located in area of quite complex topography.

@@

```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1. A) The simulation starts by drawing a single year from the posterior density estimate.  B) This then corresponds to an elevation range on the interpolated displacement curve. A single elevation is drawn uniformly from this range using 5cm intervals. C) The sea-level is then adjusted on the DTM to this elevation and the various distance measures are found. D) The numerical result of the simulation run.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
```

@@

```{r distances}
distances2 <- distances %>%
  filter(year <= -2500 & rcarb_cor == "t" & type == "forager") %>% 
  mutate(hordist = ifelse(hordist < 0, 0, hordist),
         topodist = ifelse(topodist < 0, 0, topodist),
         vertdist = ifelse(vertdist < 0, 0, vertdist))

distances3 <- distances %>%
  filter(year <= -4000 & rcarb_cor == "t" & type == "forager") %>% 
  mutate(hordist = ifelse(hordist < 0, 0, hordist),
         topodist = ifelse(topodist < 0, 0, topodist),
         vertdist = ifelse(vertdist < 0, 0, vertdist))
```
# Simulation results

Overall, as is indicated by the measures for central tendency and the almost solid line along the 0m mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)). Some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \@ref(fig:results)A), but if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \@ref(fig:results)B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the rapid RSL fall in the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level, as indicated on the plot for vertical distance. While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE. The chronological smearing following from the uncertainty in the ^14^C-dates means that while the results cannot be used to directly inform discussions that deal with the century scale around these chronological transitions [e.g.  @prescott2020; @solheim2021], the findings are nonetheless in clear agreement with the general chronological framework in the literature.

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b]. While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water. It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.

@@

# Subtract one for Gunnarsrød 5 (see text)
synchdata$meso[synchdata$cross == "false"] <- 
  synchdata$meso[synchdata$cross == "false"] - 1

@@

```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites. Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occurred. Results associated with agricultural activites are plotted in grey. The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites. The second row B) shows the result of excluding these. The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results.png"))

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
   - \usepackage[font={small}]{caption}
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
# always_allow_html: yes # Uncomment if compiling to word
output:
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed 'shoreline dating' which is widely applied. However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located below the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. Monte Carlo simulation is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous hypotheses that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

keywords: |
  Shoreline dating; Stone Age; Settlement patterns; Scandinavia; Relative sea-level change
nocite: | 
  @telford2004
---

@@

```{r libraries, eval = TRUE, results = 'hide'}
library(here)
library(tidyverse)
library(sf)
library(ggridges)
library(ggthemes)
library(ggnewscale)
library(patchwork)
library(terra)
library(raster)
library(oxcAAR)
library(vapour)
library(topoDistance)
library(IRanges)
library(stringr)
library(kableExtra)
library(english)
quickSetupOxcal()

@@

# Load preprocessed site and radiocarbon data, results, and load functions
load(here("analysis/data/derived_data/01data.RData"))
load(here("analysis/data/derived_data/02data.RData"))
load(here("analysis/data/derived_data/06data.RData"))
load(here("analysis/data/derived_data/08agediff.RData"))
load(here("analysis/data/derived_data/08posterior_rcarbon.RData"))
load(here("analysis/data/derived_data/09data.RData"))
source(here("analysis/script/04analysis_functions.R"),
       local = knitr::knit_global())

# Exclude Lunaveien, Dybdalshei 2 and Frebergsvik C (see supplementary material)
sites_sa <- filter(sites_sa, !(name %in% c("Dybdalshei 2", 
                                           "Lunaveien", "Frebergsvik C")))

# Do the same for radiocarbon dates intersecting the Stone Age.
rcarb_sa <- filter(rcarb_sa, site_name %in% unique(sites_sa$name))

@@

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit (the highest elevation of the sea after the retreat of the ice) were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [see e.g. @berg-hansen2009; @eskeland2017; @nummedal1923; @simpson2009]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @berg-hansen2022, 644; @ilves2011].

@@

Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability mass functions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @akerlund1995; @åstveit2018; @berg-hansen2022; @solheim2020; see also @bjerck2008a; @kleppe1985; @ramstad2009]. This approach has a couple of limitations. First, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g. @åstveit2018; @fossum2020; @møller1987; @persson2008; @rosenvinge2022]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that the conclusion supports the hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.

@@

The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase. The isobases are here contours indicating equal shoreline displacement over the same time span [@svendsen1987, 116]. To minimise the impact of variable uplift rates, the cored basins are located in as constrained of an area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)). Furthermore, as the uplift has been greater immediately following the retreat of the ice, such adjustments, and thus potential uncertainty, will be more critical further back in time. The resulting SLIPs are most commonly interpreted as representing the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range. For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [30cm at the standard port Helgeroa in Larvik, @mapping2021].

As the displacement curves and their trajectory are quite complex constructs and the integrated result of both expert knowledge and more objectively quantifiable parameters, the geologists that have undertaken the studies have not found reason to assign variable uncertainty within the confidence envelopes of the displacement curves [@sørensen2014, 44; @romundset2018, 187]. The reason for this is that the trajectory of the curves is not only based on radiometric dates, the uncertainty of which are well defined, but are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g. @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

The archaeological data compiled for the analysis consists of excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number `r nrow(sites)` sites, of which `r length(unique(rcarbon$site_name))` are associated with the total of `r nrow(rcarbon)` radiocarbon dates. Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon date ranges that intersect the Stone Age (9500--1700 BCE), with 95% probability. These sites and ^14^C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History of the University of Oslo---the institution responsible for archaeological excavations and data curation in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

@@

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of RSL change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year weighted by the posterior probability distribution of a given occupation phase of a site (Figure \@ref(fig:example-2)). This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDistance package for R [@wang2019]. The topographic distance was measured between the points on the site and sea polygons that were identified as being the closest when measured horizontally. Because it is measured as the shortest topographic path between the horizontally closest points, this means that the distance does not necessarily match the closest topographic distance if the entirety of the polygons had been considered. Not finding the topographically closest points significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact on the results, given the distances considered. The shortest topographic path was found using the Moore neighbourhood of eight cells [e.g. @conolly2006, 253; @herzog2013].

In the case where the sea polygons intersect the site polygon, all distance measures were set to zero. In the case that the sea polygons completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values were therefore retained with the exception of the sites of Gunnarsrød 5 and Pjonkerød R1, where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

This process was repeated 1000 times for each phase for each site (Figure \@ref(fig:example-3)). The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5 [cf. @crema2010, 1125]. This evaluation is presented in the supplementary material. Hovland 5 was chosen for this assessment as it has an imprecise age and is located in area of quite complex topography.

@@

```{r example-2, echo = FALSE, fig.cap = "Example of a single simulation run on the site Hegna vest 1. A) The simulation starts by drawing a single year, weighted by the the posterior probability distribution.  B) This then corresponds to an elevation range on the interpolated displacement curve. A single elevation is drawn uniformly from this range using 5cm intervals. C) The sea-level is then adjusted on the DTM to this elevation and the various distance measures are found. D) The numerical result of the simulation run.", out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/example_sim_map.png"))
```

@@

```{r distances}
# Various subsets for which some statistics are reported in the text.

distcor <- filter(distances, rcarb_cor == "t")

# Results older than 7500 BCE
r1 <- distcor %>% filter(year < -7500)

# Statistics for vertical distance
rs1 <- r1 %>% 
  summarise(median = median(vertdist),
          max = max(vertdist),
          min = min(vertdist))

# Number of sites with results higher than 5m
r1a5 <- r1 %>% group_by(sitename) %>% summarise(min = min(vertdist)) %>%
  filter(min < 5) %>% nrow()

# Results older than 2500 BCE
r2 <- distcor %>% filter(year > -2500)

# Statistics
rs2 <- r2 %>% 
  summarise(
    mediandist = median(vertdist),
    maxdist = max(vertdist)
  )

# Number of sites with results that include a vertical distance of zero
r2a0 <- r2 %>% group_by(sitename) %>% summarise(min = min(vertdist)) %>%
  filter(!(min > 0)) %>% nrow()

# Results between 4000 BCE and 2500 BCE
r3 <- distcor %>% filter(year < -2500 & year > -4000)

# Statistics
rs3 <- r3 %>% 
  summarise(median = median(vertdist))

# Number of sites with results higher than 25m
r3a25 <- r3 %>% group_by(sitename) %>% summarise(min = min(vertdist)) %>% 
  filter(min > 25) %>% nrow()

# Data when setting negative results to zero, to be used in the text
distances2 <- distances %>%
  filter(year <= -2500 & rcarb_cor == "t" & type == "forager") %>% 
  mutate(hordist = ifelse(hordist < 0, 0, hordist),
         topodist = ifelse(topodist < 0, 0, topodist),
         vertdist = ifelse(vertdist < 0, 0, vertdist))

distances3 <- distances %>%
  filter(year <= -4000 & rcarb_cor == "t" & type == "forager") %>% 
  mutate(hordist = ifelse(hordist < 0, 0, hordist),
         topodist = ifelse(topodist < 0, 0, topodist),
         vertdist = ifelse(vertdist < 0, 0, vertdist))

```
# Simulation results

Overall, as is indicated by the measures for central tendency and the almost solid line along the 0m mark on the y-axes, the simulations show that the sites tend to have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)). As is also illustrated by the measures for dispersion, some of the sites are situated considerable distances from the shoreline when the dates believed to be erroneous in the original reports are included (Figure \@ref(fig:results)A). However, if one accepts the interpretation that these do not date the main occupation of the sites, as is indicated by the artefact inventories, Figure \@ref(fig:results)B gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. The distances for the earliest sites appears somewhat high, with the highest vertical distance of the results older than 7500 BCE being `r round(rs1$max, 1)`m. But this can likely be explained as the result of the rapid RSL fall in the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty of the ^14^C-dates to give a wider possible elevation range for the simulated sea-level. This is also indicated by the fact that the median vertical distance for the same simulation results is `r round(rs1$median, 1)`m, and `r r1a5` of the `r length(unique(r1$sitename))` sites associated with these results have simulated vertical distances that extend below 5m.

Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. After around 2500 BCE several sites are situated a considerable distance from the reconstructed shoreline, and while a couple remain horizontally and topographically close, most appear to be considerably elevated above the sea-level. Of the results from after 2500 BCE, which are associated with `r length(unique(r2$sitename))` sites, only `r as.english(r2a0)` has simulation results for vertical distance that includes zero. The highest simulated vertical distance among these is `r round(rs2$max, 1)`m and the median is `r round(rs2$median, 1)`m. While the sample size is limited, there are also a couple of sites located some distance from the shoreline just after 4000 BCE. Of the `r length(unique(r3$sitename))` sites associated with the period between 4000 and 2500 BCE, `r r3a25` sites have all vertical distance results above 25m. However, the median vertical distance of the results from this period is only `r round(rs3$median, 1)`m, indicating that while some sites have a markedly withdrawn location, most are still situated close to the shoreline. The chronological smearing following from the uncertainty in the ^14^C-dates means that while the results cannot be used to directly inform discussions that deal with the century scale around these chronological transitions [e.g.  @prescott2020; @solheim2021], the findings are nonetheless in clear agreement with the general chronological developments suggested in the literature.

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. @berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites. These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022]. While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent an upper constraining limit for the sea-level, as they would not have been in use when located under water. It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.

@@

# Number of site phases for all dates and only dates corresponding to site
# inventories.
nphases <- ardates %>%  group_by(site_name) %>% distinct(group) %>% nrow()
nphasescor <- rdates %>% group_by(site_name) %>% distinct(group) %>% nrow()

# Number of sites for simulation results older than 2500 BCE
nsitesdist25 <- rdates %>% filter(dates < -2500 & probabilities > 0) %>%
  group_by(site_name) %>% distinct(site_name) %>% nrow()

# Number of sites for simulation results older than 4000 BCE
nsitesdist4 <- rdates %>% filter(dates < -4000 & probabilities > 0) %>%
  group_by(site_name) %>% distinct(site_name) %>% nrow()

# Number of site phases for simulation results older than 2500 BCE
nphasesdist25 <- rdates %>% filter(dates < -2500 & probabilities > 0) %>%
  group_by(site_name) %>% distinct(group) %>% nrow()
# Number of site phases for simulation results older than 4000 BCE
nphasesdist4 <- rdates %>% filter(dates < -4000 & probabilities > 0) %>%
  group_by(site_name) %>% distinct(group) %>% nrow()

# Correspondence between shoreline and radiocarbon dates.
# Subtract one for Gunnarsrød 5 (see text)
synchdata$meso[synchdata$cross == "false"] <- 
  synchdata$meso[synchdata$cross == "false"] - 1

@@

```{r results, echo = FALSE, fig.cap = paste0("The result of running the analysis across all sites. Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occurred. Results associated with agricultural activities are plotted in grey. The first row A) shows the result of including all dates to the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites (", length(unique(distances$sitename)) ," sites and ", nphases ," site phases). The second row B) shows the result of excluding these (resulting in ", length(unique(distances[distances$rcarb_cor == "t",]$sitename)), " sites and ", nphasescor ," site phases). The table under each plot lists some corresponding statistics for central tendency and dispersion."), out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results.png"))
